Cheminformatics: The Computer Science of Chemical Discovery

### 2011-08-02 paper.tex
Added in Joergs edits

\category{H.4}{Information Systems Applications}{Miscellaneous}
\terms{
\begin{itemize}
\item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H$^+$) to an atom, molecule, or ion. The different states (different number of H$^+$ or on different atoms) are called protonation states for a molecular compound (e.g. de-protonated, mono-protonated, di-protonated, or any variations of those states).
\item \textit{Bioassay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
\item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is usually restricted to small molecules (chemical compounds).
\item \textit{Mass spectrometry} - Mass spectrometry (MS) is an analytical technique that measures the mass-to-charge ratio of charged particles. It is used for determining masses of particles, for determining the elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such as peptides and other chemical compounds.
\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions.
\item \textit{Ligand (biochemistry)} - a substance that binds to a protein.
\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance.
\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework.
\item \textit{Biodegradation} - is the chemical dissolution of materials by bacteria or other biological means.
\item \textit{Reaction catalysis} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
\item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single bonds. Single possible states of compounds are called conformers.
\item \textit{SMARTS} - SMiles ARbitrary Target Specification \\(SMARTS) is a language for specifying substructure patterns in molecules
\end{itemize}
}
\keywords{cheminformatics, graphs, chemistry, databases, knowledge management, open source, open data}

@@

Most cheminformatics applications rely on large databases of chemical structures
and their properties. Organizing and maintaining such databases, searching based
on features of chemical structures, and clustering similar structures together,
are essential utilities to enable many scientific applications. Yet, each one of
those areas poses its unique computer science challenges.

@@

However, the future still leaves many challenges ahead. Using external/public databases
with chemical and bioactivity data remains a challenge due to differences in
ontologies, synchronization and maintenance, and efficient
substructure and similarity search within complex data types
(compounds). A
collaboration with external parties, e.g. contract research, poses other
problems with duplicate checking for compounds (e.g. tautomeric or salt
structures) and efficient data synchronization, and most often a double-registration is
required for fulfilling the data quality processes on both sides. If external
partners are small they often do not even have sufficient IT resources
themselves and thus rely on external services. Cloud services could help
not only to provide a service infrastructure for all parties involved,
but also to provide the required private/public access toolbox. Furthermore,
there are still many data sources not being indexed properly, since chemistry
patents are often cryptic and image/text mining remains a puzzle. One could
argue that the image/text mining in scientific journals is less challenging; in
fact, it remains a huge challenge and many chemists do not trust automated
non-expert validated mining sources. The closed CAS (Chemical Abstract Service)
is a highly trusted source, and public chemistry and bioactivity space has to
improve quality and interconnectivity to compare with it. Efforts like ChEMBL and PubChem-BioAssay
are on the right track. Still, improving data quality and standards between
public and closed sources will be absolutely critical for ensuring constant
growth, usage, and collaboration scenarios between private and public parties.
Only when the technical and scientific questions are solved can we ensure that
legal questions about contribution metrics will get enough room to grow.

>>>>

\category{H.4}{Information Systems Applications}{Miscellaneous}
\terms{
\begin{itemize}
\item \textit{Bioassay} - A system measuring the biological activity of a chemical compound on a particular biological assay (with all its parameters).
\item \textit{Biodegradation} - is the chemical dissolution of materials by bacteria or other biological means.
\item \textit{Conformation and conformational isomerism} - In chemistry, conformational isomerism is a form of stereoisomerism in which the isomers can be interconverted exclusively by rotations about formally single bonds. Single possible states of compounds are called conformers.
\item \textit{Ligand (biochemistry)} - a substance that binds to a protein.
\item \textit{Ligand (coordination chemistry)} - an atom, ion, or functional group that donates one or more of its electrons through a coordinate covalent bond to one or more central atoms or ions.
\item \textit{Mass spectrometry} - Mass spectrometry (MS) is an analytical technique that measures the mass-to-charge ratio of charged particles. It is used for determining masses of particles, for determining the elemental composition of a sample or molecule, and for elucidating the chemical structures of molecules, such as peptides and other chemical compounds.
\item \textit{Metabolites} - Metabolites are the intermediates and products of metabolism. The term metabolite is usually restricted to small molecules (chemical compounds).
\item \textit{Proteochemometric modeling} - combining data mining for chemical compounds (ligands) and proteins within the same machine learning framework.
\item \textit{Protonation and protonation state} - In chemistry, protonation is the addition of a proton (H$^+$) to an atom, molecule, or ion. The different states (different number of H$^+$ or on different atoms) are called protonation states for a molecular compound (e.g. de-protonated, mono-protonated, di-protonated, or any variations of those states).
\item \textit{Reaction catalysis} - Catalysis is the change in rate of a chemical reaction due to the participation of a substance called a catalyst.
\item \textit{Receptor (biochemistry)} - in biochemistry, a protein molecule that receives and responds to a neurotransmitter, or other substance.
\item \textit{Regulation authorities} - In the US the Food and Drug Administration (FDA) is responsible for protecting and promoting public health through the regulation and supervision of e.g. dietary supplements, prescription and over-the-counter pharmaceutical drugs (medications), vaccines, biopharmaceuticals, medical devices, and many other health related items. The European Medicines Agency (EMA) is a European agency for the evaluation of medicinal products.
\item \textit{SMARTS} - SMiles ARbitrary Target Specification \\(SMARTS) is a language for specifying substructure patterns in molecules
\item \textit{Stereoisomer/stereoisomerism} - Stereoisomers are isomeric molecules that have the same molecular formula and sequence of bonded atoms (constitution), but that differ only in the three-dimensional orientations of their atoms in space.
\item \textit{Tautomerism} - Tautomers are isomers (same molecular formula but different structural formula) of organic compounds that readily interconvert by a chemical reaction called tautomerization.
\end{itemize}
}
\keywords{cheminformatics, graphs, chemistry, databases, knowledge management, open source, open data}

@@

Most cheminformatics applications rely on large databases of chemical structures, 
their properties, and relationships to e.g. biological targets.
Organizing and maintaining such databases, searching based
on features of chemical structures, and clustering similar structures together,
are essential utilities to enable many scientific applications. Yet, each one of
those areas poses its unique computer science challenges.

@@

However, the future still leaves many challenges ahead. Using external/public databases
with chemical and bioactivity data remains a challenge due to differences in
ontologies, synchronization and maintenance, and efficient
substructure and similarity search within complex data types
(compounds). A
collaboration with external parties, e.g. contract research, poses other
problems with duplicate checking for compounds (e.g. tautomeric or salt
structures) and efficient data synchronization, and most often a double-registration is
required for fulfilling the data quality processes on both sides. If external
partners are small they often do not even have sufficient IT resources
themselves and thus rely on external services. Cloud services could help
not only to provide a service infrastructure for all parties involved,
but also to provide the required private/public access toolbox. Furthermore,
there are still many data sources not being indexed properly, since chemistry
patents are often cryptic and image/text mining remains a puzzle. One could
argue that the image/text mining in scientific journals is less challenging; in
fact, it remains a huge challenge and many chemists do not trust automated
non-expert validated mining sources. The closed CAS (Chemical Abstract Service)
is a highly trusted source, and public chemistry and bioactivity space has to
improve quality and interconnectivity to compare with it. Efforts like ChEMBL and PubChem-BioAssay
are on the right track. Still, improving data quality and standards between
public and closed sources will be absolutely critical for ensuring constant
growth, usage, and collaboration scenarios between private and public parties.
In a business context we need to address ROI (return-of-investment) questions, not only for single
institutions and companies, but also for collaboration networks of legal partners.
So, only when the technical and scientific questions are solved can we ensure that
legal questions about contribution metrics will get enough room to grow. Solving challenges is one thing (science and technical),
ensuring that the contributing partners get rewarded by their contribution another (legal).
